The completion of this manuscript was made possible through the invaluable contributions of a number of people. First, I would like to express my gratitude to my supervisors Mark Braverman and Avner Magen. They both provided me with a deep perspective of theoretical computer science and guidance on many technical aspects of this thesis. Their inspiration has been truly invaluable to me. Moreover, I would like to thank Allan Borodin, Toni Pitassi and Mike Molloy for agreeing to serve as members of my departmental examination committee. Moreover, I would like to sincerely thank Joel A. Tropp for his careful assessment of our work and for providing useful suggestions for improvements.
%

%
During my time in graduate school, I had the great pleasure to collaborate with many researchers. I would like to thank all my co-authors: Haim Avron, Christos Boutsidis, Petros Drineas, Jeff Edmonds, Nick Freris, Piotr Indyk, Pascal Koiran, Avner Magen, Mike Mahoney, Tassos Sidiropoulos, Sivan Toledo and Michail Vlachos. I truly enjoyed collaborating with them and I have learned a lot from each one of them.
%

%
The department of computer science at the university of Toronto (UofT) is a great academic environment with many great scholars. In particular, I would like to thank all the members of the theory group. I have also learned a lot from fellow graduate students and postdocs especially Aki, Arkadev, Brendan, Costis, Dai, Eden, George, Jeremy, Joel, Kaveh, Lila,  Natan, Per, Periklis, Siavosh, Tassos, Wesley, Yevgeniy, Yu and Yuval.
%

%
Many thanks go to IBM Research Zurich for hosting me as a research intern during the summer and fall of 2011. The initial discussions about the randomized gossip algorithms of Chapter~\ref{chap:graph} were initiated there. In addition, I would like to thank the department of computer science at Princeton University for hosting me as a visiting student during the winter of 2012 (especially Mark for setting up everything).
%

%
I would like to thank all Greek graduate students at UofT (a.k.a. Greek mafia or ``grspamites'') for all the moments of joy and fun that we shared together. Last but not least, I would like to thank Maria for her support, encouragement, and for making my journey through graduate school, pleasant and full of beautiful moments.
%

%
Finally, my heartfelt thanks go out to my family and friends in Greece for their constant support and encouragement.